---
title: "Waldo examples"
output: github_document
---

```{r setup, include=FALSE}

if(!require("waldo")) {
  install.packages("waldo")
  library(waldo)
}

library(tidyverse)

```

# Introduction

[Waldo](https://www.tidyverse.org/blog/2020/10/waldo/) was originally created as a tool for testing equality of objects when developing R packages [(see `testthat::expect_equal()`)](https://testthat.r-lib.org/reference/equality-expectations.html). The purpose of the main function [`compare()`](https://rdrr.io/cran/waldo/man/compare.html) is to determine the difference between two R objects and display them. That way, if there are differences, the test will fail and one can determine the source of the bug. However, could we not use this function to ascertain differences in data sets?

## Metadata

A common problem is that we have a matrix of say, KEGG ortholog counts, and a data-frame full of samples. NB: This is example data from a mouse project.

```{r}

my_data <- read_tsv("data_matrix.tsv", show_col_types = FALSE)
my_samples <- read_tsv("sample_sheet.tsv", show_col_types = FALSE)

```

### Subsetting

A common practice is to subset the data to "drill-down" to pertinent details. We might filter out the controls or just look at the top 10 genes in a dataset.

Like so:

```{r}

my_subset <- my_data %>%
  group_by(geneID) %>%
  summarise(sum_count = sum(count)) %>%
  slice(1:10) %>%
  left_join(my_data, by = "geneID") %>%
  select(SampleID, count, geneID)

my_subset %>%
  ggplot(aes(x = SampleID, y = count, fill = geneID)) +
  geom_bar(stat = "identity", position = "fill") +
  theme_bw() +
  theme(axis.text.x = element_blank(), strip.background = element_blank()) +
  scale_fill_brewer(palette = "Paired") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Top 10 KEGG orthologs",
       y = "Proportion among the top 10",
       x = "")

```

## Which ones were left out?

With Waldo, we can quickly determine which genes were left out of the graph:

```{r}

waldo::compare(unique(my_data$geneID), unique(my_subset$geneID))

# compare(my_data, my_subset)
# can not directly compare dataframes!
# gives error: Error in compare_proxy.spec_tbl_df(x, paths[[1]]) : unused argument (paths[[1]])

```

Another example with metadata this time. Subsetting samples sometimes gets us into trouble if we have also subsetted a matrix of counts.

```{r}

sample_subset1 <- my_samples %>%
  filter(study_group %in% "Blue")

my_matrix <- my_data %>%
  filter(SampleID %in% sample_subset1$SampleID) %>%
  pivot_wider(names_from = SampleID, values_from = count) %>%
  column_to_rownames(var = "geneID") %>%
  as.matrix()

head(my_matrix)

```

```{r}

sample_subset2 <- my_samples %>%
  filter(study_group %in% "Yellow")

my_error <- try(my_matrix[,sample_subset2$SampleID])

```

Again, you can quickly find which samples changed:

```{r}

compare(colnames(my_matrix), sample_subset2$SampleID, max_diffs = Inf)

```

